---
title: "Top-down disaggregation of Burkina Faso 2019 census - Scripts"
author: "Edith Darin, WorldPop"
always_allow_html: true
header-includes:
   - \usepackage{amsmath}
output:
  github_document:
    toc: true
geometry: margin=2.1cm
documentclass: article
bibliography: [book.bib]
biblio-style: apalike
---

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/ecd1u18/Documents/BFA_population_v1_0_methods" )
knitr::opts_chunk$set(message = FALSE, echo = T, warning = FALSE, eval = FALSE)
root_path <-  "C:/Users/ecd1u18/Documents/BFA_population_v1_0_methods/scripts"
data_path <- "C:/Users/ecd1u18/Documents/BFA_population_v1_0_methods/data"
output_path <-"//worldpop.files.soton.ac.uk/worldpop/Projects/WP517763_GRID3/Working/BFA/model"
```

# Introduction

This document presents the data and R code used to estimate the top down model for disaggregating 2019 Burkina Faso census totals [@institutnationaldelastatistiqueetdeladémographie2019]. It is mainly based on scripts developed by Bondarenko for the WorldPop project [@bondarenko2018].

Please note that we used only their scripts aiming at estimating the model. All the pre-processing of the data input was done directly in our custom scripts as well as the prediction.

The data can be found here [TO BE COMPLETED WHEN RELEASED]. Supporting scripts are in the `scripts`folder.

# Environment setup

We had first to configure the setting specifically for Burkina Faso as specified in Bondarenko [@bondarenko2018], that can be seen in the `input_BFA.R` script.

```{r}
source(paste0(root_path,"/input_BFA.R"))

```

Then we need to load all the required functions for estimating the tRandom Forest model for population modelling.

```{r setup}
source(paste0(root_path,"/config.R"))
source(paste0(root_path,"/load_Packages.R"))
source(paste0(root_path,"/internal_functions.R"))  
source(paste0(root_path,"/create_dirs_for_prj.R"))
source(paste0(root_path,"/rf_functions.R")) 

if (!load.Packages())
  stop("There was an error when loading R packages")


```

We create all the necessary directories for the estimation.

```{r, echo=F}
glPaths <- create_dirs_for_prj(paste0(output_path, "/topdown/"))  

##  Get the paths to the countries' data:
rfg.data.path.countries <- glPaths$data
##  Declare where we are outputting things:
rfg.output.path.countries <- glPaths$output
##  Declare where we are outputting things:
rfg.output.path.countries.cvr <- glPaths$data_cvr
##  Declare where our temporary path is:
rfg.output.path.countries.tmp <- paste0(rfg.output.path.countries, "tmp/")
##  Retrieve the country tag:
rfg.countries.tag <- glPaths$countries_tag


##  Remove unnecessary items:
rm(glPaths)

```

# Training the model

We load the data. The data are in a table format with each row being an admin 3.

```{r}
source(paste0(root_path,"/variable_names.R")) 

train <- readRDS(paste0(data_path, "/BFA_population_v1_0_train.rds"))
predict <- readRDS(paste0(data_path, "/BFA_population_v1_0_predict.rds"))
age_sex <- read.csv(paste0(data_path, "/BFA_population_v1_0_agesex.csv"), stringsAsFactors = F)

# mastergrid to convert back the prediction into raster format
masterGrid <- raster(paste0(data_path, "/BFA_population_v1_0_mastergrid.tif"))

```

We prepare the predictors and the response variable.

```{r}

# Predictors
cov_names <- colnames(train)
cov_names <- cov_names[5:(length(cov_names)-19)]
cov_names <- cov_names[!grepl("settled_binary_V3|settledArea_V3|Img_date", cov_names)]

x_data <- train[,cov_names]

# Response variable: population per sum of settled pixel size

y_data <- log(as.numeric(train$pop_density_pixels))

```

We train the model.

```{r}
## Fit the RF, removing any covariates which are not important to the model:
popfit <- get_popfit()

##  Fit the final RF 
popfit_final <- get_popfit_final()
```

# Predicting the population

We prepare the prediction data set in a table format with each row being a settled pixel.

```{r}
cov_names <- names(popfit_final$forest$xlevels)
cov_predict <- predict[, ..cov_names]
cov_predict$admin3_id <- predict$admin3_id
cov_predict$masterGrid_id <- predict$masterGrid_id
cov_predict$settledArea <- predict$settledArea_V3

```

We define the function that predict density at pixel level and convert it to a weight which is mutliply by the total census count of the related admin 3.

```{r}
predict_weights <- function(df, census, model=popfit_final){
  prediction_set <- predict(model, 
                            newdata=df, 
                            predict.all=TRUE)
  output <- data.table(rf_pred = exp(apply(prediction_set$individual, MARGIN=1, mean)))
  
  output$admin3_id <- df$admin3_id
  output$masterGrid_id <- df$masterGrid_id
  output$weight <- output$rf_pred
  output$pop <- (output$weight/sum(output$weight))*census
  
  fwrite(output, paste0(output_path, "/topdown/output/prediction/predictions_",
                        df$admin3_id[1], ".csv"))
}

```

We run the prediction in parallel mode, each batch containing all the pixels of a given admin 3;

```{r}
cov_predict_list <- as.data.frame(cov_predict) %>% 
  group_by(admin3_id) %>% 
  group_split()

co <- detectCores()-2
tic()
cl <- makeCluster(co)
registerDoParallel(cl)
predicted <- NULL
predicted <- foreach(
  i=1:length(cov_predict_list), 
  .packages=c("tidyverse", "data.table", "randomForest")) %dopar% {
    predict_weights(
      cov_predict_list[[i]],
      master %>% filter(admin3_id==i) %>% dplyr::select(population) %>% unlist()
    )
  } 

stopCluster(cl)
toc() #30sec


```

# Gridding the predicted pop

We convert the prediction tables into a raster representing the gridded population.

```{r}
# reading in the predictions
predictions_list <- list.files(
  paste0(output_path, "/topdown/output/prediction/"), pattern = ".csv") 
print(length(predictions_list))

readPredictions <- function(file){
  df <- fread(paste0(output_path,  "/topdown/output/prediction/", file)
  )
  return(df)
}

tic()
predictions <- do.call("rbind", 
                       lapply(
                         predictions_list, 
                         function(x) readPredictions(x))
)
toc() #20

# assign the prediction to its pixel using the mastergrid id

raster_df <- data.table(masterGrid[])
names(raster_df) <- "masterGrid_id"

raster_df <- predictions[raster_df, on="masterGrid_id"]
r <- masterGrid
r[] <-  raster_df$pop

# writing the output raster

writeRaster(r, 
            paste0(output_path, "/BFA_population_v1_0_gridded.tif"),
            overwrite=T)
```

# Disaggregating by age and sex groups

To obtain age and sex disaggregation at pixel level, we multiply the predicted population count by the national age and sex proportions.

```{r}
for(i in 1:(ncol(age_sex)-1)){
  prop <- age_sex[1,i] %>% unlist()
  print(names(prop))
  r_decomp <- r* prop
  writeRaster(r_decomp, 
              paste0(output_path, 
                     'BFA_population_v1_0_agesex/BFA_population_v1_0_agesex_',
                     names(prop),
                     '.tif'),
              overwrite=T)
}
```

# References
